---
title: "final"
output: html_document
---

Link to GitHub repo: https://github.com/eliana427/ENVS-193DS_spring-2025_final.git
```{r}
library(tidyverse)
library(janitor)
sst <- read_csv("data/SST_update2023.csv")
```

Problem 1. Research writing

a. Transparent statistical methods

In part one, my co-worker likely did a Pearson correlation (if each variable is normally distributed), but if they're not normally distributed than the test was likely a Spearman rank correlation. In part 2, my co-worker likely did a one way ANOVA test (for normally distributed variables) or a Kurskal-Wallis test (data is not normally distributed).

b.  More information needed

The effect size (in this case, $\eta^2$) would be helpful because it gives information about how big the difference is between groups, specifically how much the type of source impacts the variation of average nitrogen load. A post hoc Tukey's HSD test would be helpful because it provides information about which groups in particular differ in nitrogen load by comparing the pairs of sources directly.

c. Suggestions for rewriting

The average load of nitrogen (kg/year) differed a lot between tested sources (urban land, atmospheric deposition, fertilizer, waste water treatment, and grasslands) (one way ANOVA...). Effect size (). Tukey's HSD (). **finsih later**


Problem 2. Data visualization

```{r}
sst_clean <- sst |> 
  clean_names() |> 
  group_by(date) |> 
  summarize(
    mean_sst = mean(mean_monthly_sst, na.rm = TRUE),
    sd_sst = sd(mean_monthly_sst, na.rm = TRUE),
    n = n()
  )

sst_clean

```

