---
title: "Final - ENVS 193DS"
author: "Eliana Shandalov"
date: "June 10, 2025"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

Link to GitHub repo: https://github.com/eliana427/ENVS-193DS_spring-2025_final.git
```{r}
# reading in packages
library(tidyverse)
library(janitor)
library(lubridate)
library(here)
library(scales)
library(ggeffects)
library(MuMIn)
library(DHARMa)

# reading in data sets
sst <- read_csv(
  here("data", "SST_update2023.csv"))
nest_boxes <- read_csv(
  here("data", "occdist.csv"))
```

# Problem 1. Research writing

## a. Transparent statistical methods

In part one, my co-worker likely did a Pearson correlation (if each variable is normally distributed), but if they're not normally distributed than the test was likely a Spearman rank correlation. In part 2, my co-worker likely did a one way ANOVA test (for normally distributed variable and sources have equal variances) or a Kurskal-Wallis test (data is not normally distributed).

## b.  More information needed

The effect size (in this case, $\eta^2$) would be helpful because it gives information about how big the difference is between groups, specifically how much the type of source impacts the variation of average nitrogen load. A post hoc Tukey's HSD test would be helpful because it provides information about which groups in particular differ in nitrogen load by comparing the pairs of sources directly.

## c. Suggestions for rewriting

We found a large difference ($\eta^2$ = effect size) in average load of nitrogen (kg/year) between tested sources (urban land, atmospheric deposition, fertilizer, waste water treatment, and grasslands) (one-way ANOVA, F(4, df within groups) = F value, p = 0.03, $\alpha$ = significance level)). A post hoc comparison using Tukey's HSD indicated that there were differences in the mean load of nitrogen between two of the of sources (amount kg/year less, x% CI: [lower bound, upper bound]) and another two of the of sources (amount less kg/year, x% CI: [lower bound, upper bound]).


# Problem 2. Data visualization

## a. Cleaning and summarizing
```{r}
sst_clean <- sst |> # store sst as new object called sst_clean
  clean_names() # make all names lowercase, and if there were spaces then change to underscores

sst_clean <- sst_clean |> # assigns sst_clean to itself to use in next functions
  mutate(
    date = as.Date(date), # make sure that R reads this column as a date that can be separated
    month = month(date, label = TRUE, abbr = FALSE), # separate out months from date column
    year_extracted = year(date) # extract the year from the data set
  ) |> 
  filter(year_extracted >= 2018) |> # filter the data set to only include dates 2018 and later
  mutate(year = factor(year_extracted), # read year as a factor
    date = NULL, # remove date column from displayed data in console
    site = NULL, # remove site column from displayed data in console
    latitude = NULL, # remove latitude column from displayed data in console
    longitude = NULL # remove longitude column from displayed data in console
  ) |> 
  group_by(year, month) |> # group by year and month
  summarize(mean_monthly_sst = mean(temp, na.rm = TRUE)) |> # calculate mean temperature per group
  ungroup() |>  # removes groupings back to original data frame

# rename all months in month column to abbreviations  
mutate(month = case_when(
  month == "January" ~ "Jan",
  month == "February" ~ "Feb",
  month == "March" ~ "Mar",
  month == "April" ~ "Apr",
  month == "May" ~ "May",
  month == "June" ~ "Jun",
  month == "July" ~ "Jul",
  month == "August" ~ "Aug",
  month == "September" ~ "Sep",
  month == "October" ~ "Oct",
  month == "November" ~ "Nov",
  month == "December" ~ "Dec"
)) |> 
  
mutate(month = factor(month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
ordered = TRUE)) # sort months in order when they appear on graph

slice_sample(sst_clean, n = 5) # display five random rows of the data set 

str(sst_clean) # show structure of the data set
```

## b. Visualize the data

```{r}
sst_plot <- sst_clean # creating new object for plot

ggplot(data = sst_clean, # use data from sst_clean
       aes(x = month, # x axis is month
           y = mean_monthly_sst, # y axis is mean_monthly_sst
           color = year, # color will be based on year
           group = year)) + # data will be grouped by year+ 
geom_line() + # create a line plot
  scale_color_manual(values= c("#2E004F", "#4B0082", "#6A0DAD", "#8A2BE2", "#A24AC3", "#BF65D9")) + # colors will be a gradient with 2018 being the darkest and 2023 being the lightest
geom_point(size = 1.1) + # add points to the lines of size 1.1
  labs(x = "Month", # rename the x axis
       y = "Mean monthly sea surface temperature (°C)", # rename the y axis
       color = "Year") + # rename the legend which is based on color
  theme(panel.grid = element_blank(), # remove the grid lines
        panel.background = element_blank(), # make panel background blank
        panel.border = element_rect(color = "grey24", fill = NA, linewidth = 0.5), # add border to plot of width 0.5 but don't fill it in
        axis.ticks.length.x = unit (1, "mm"), # make axis ticks on x axis 1 mm
        axis.ticks.length.y = unit (1, "mm"), # make axis ticks on y axis 1 mm
        legend.position = "inside", # place legend inside of plot
        legend.position.inside = c(0.15, 0.7)) # legend will be positioned at these coordinates

```

# Problem 3. Data analysis

## a. Response variable

The 0's and 1's represent if the nest box is occupied or not by a particular species; a 0 means means not occupied by these species, and a 1 means occupied by this species. The "empty" column with a 1 means the box was empty, and with a 0 means it had some species living in it.

## b. Purpose of study

Swift Parrots are far more selective about where they nest compared to Tree Martins and Common Starlings, so the nest boxes are designed for them but are at risk of occupancy of these competitors. Swift Parrots are subject to a higher risk of extinction because of human impacts on forests, more so than Tree Martins and Common Starlings because they are less selective than Swift Parrots.

## c. Difference in "seasons"

The authors of the study explain that the "seasons" are the years 2016 and 2019, which were important to test how effective the use of permanent boxes are. In 2016, the researchers put new nest boxes at a breeding sites of Swift Parrots, where a mast tree flowering event had occurred and the boxes remained in place when in 2019 another mast tree flowering event occurred.

## d. Table of models

| Model number | Season | Distance to Forest Edge | Model Description                 |  
|:------------:|:------:|:-----------------------:|-----------------------------------|  
| 1            |        |                         | no predictors (null model)        | 
| 2            |    X   |           X             | all predictors (saturated model)  |   
| 3            |        |           X             | distance to forest edge           |    
| 4            |    X   |                         | season                            |    

```{r message=FALSE, echo=FALSE}
nest_boxes_clean <- nest_boxes |> # store sst as new object called sst_clean
  clean_names() |> # make all names lowercase, and if there were spaces then change to underscores
  arrange(season, edge_distance)
```

## e. Run the models

```{r}
# model 1, null model: where any species is present
model1 <- glm(sp ~ 1, # formula for null model
                data = nest_boxes_clean, # use nest_boxes_clean data frame
              family = "binomial") # this is a binomial distribution

# model 2, saturated model: where edge distance and season impact whether any species is present
model2 <- glm(sp ~ edge_distance + season, # formula for saturated model
                data = nest_boxes_clean, # use nest_boxes_clean data frame
              family = "binomial") # this is a binomial distribution

# model 3, one predictor: where edge distance impacts if swift parrots are present
model3 <- glm(sp ~ edge_distance, # formula for using edge_distance as only predictor
                data = nest_boxes_clean, # use nest_boxes_clean data frame
              family = "binomial") # this is a binomial distribution

# model 4, other predictor: where season impacts if swift parrots are present
model4 <- glm(sp ~ season, # formula for using season as only predictor
                data = nest_boxes_clean, # use nest_boxes_clean data frame
              family = "binomial") # this is a binomial distribution

```

## f. Check the diagnostics

```{r}

# show model 1 on simulated residual plot
plot(
  simulateResiduals(model1)
)

# show model 2 on simulated residual plot
plot(
  simulateResiduals(model2)
)

# show model 3 on simulated residual plot
plot(
  simulateResiduals(model3)
)

# show model 4 on simulated residual plot
plot(
  simulateResiduals(model4)
)
```

## g. Select the best model

```{r}
# run AIC comparison to find out which model is best to represent data
AICc(model1,
     model2,
     model3,
     model4) |> 
  arrange(AICc)
```

The best model as determined by Akaike's Information Criterion (AIC) is the one that compares if Swift Parrots are present (response variable) depending on distance to forest edge and season (saturated model, all predictor variables included).

## h. Visualize the model predictions
```{r}
# model predictions for each model
par(mfrow = c(2,2)) # setting parameters, divide the area into a 2x2 grid
plot(model1, which = 1, main = "Model 1") # create diagnostic plot for model 1 and title the plot "Model 1"
plot(model2, which = 1, main = "Model 2") # create diagnostic plot for model 2 and title the plot "Model 2"
plot(model3, which = 1, main = "Model 3") # create diagnostic plot for model 3 and title the plot "Model 3"
plot(model4, which = 1, main = "Model 4") # create diagnostic plot for model 4 and title the plot "Model 4"

```

```{r}
mod_preds <- ggpredict(model2, # creating a new object for the predictions of values based on model2
                        terms = c("edge_distance [7:1072] by = 1", "season")) # the prediction will be of a distance from 7 to 1072 meters by 1m increments, and repeat again for season

mod_preds_plot <- ggplot() + # create a new object for the plot using ggplot

  geom_point(data = nest_boxes_clean, # use nest_boxes_clean data frame to add in real observations
             aes(x = edge_distance, # x axis will be edge distance
                 y = sp), # y axis will be presence of swift parrots
             size = 3, # size will be 3
             alpha = 0.4, # level of see-through the points will be
             color = "blue") + # make the points blue

  geom_ribbon(data = mod_preds, # add ribbon around lines to show confidence intervals
              aes(x = x, # x is the edge distance
                  ymin = conf.low, # lower bound CI
                  ymax = conf.high, # upper bound CI
                  fill = group), # fill color by group (season)
              alpha = 0.4) + # see through-ness of the ribbon

  geom_line(data = mod_preds, # adding predicted lines
            aes(x = x, # x is edge distance
                y = predicted, # y is predicted probability
                color = group), # color is the season
            linewidth = 1.2) + # manually setting line thickness

  scale_color_manual(values = c("2016" = "orange", "2019" = "purple")) + # 2016 line will be orange, 2019 line will be purple
  scale_fill_manual(values = c("2016" = "orange", "2019" = "purple")) + # fill colors will be orange and purple (ribbons will match lines)

  scale_y_continuous(limits = c(0, 1), # y axis will only go from 0 to 1
                     breaks = c(0, 1)) + # only have tick marks at 0 and 1
  
  theme( # customize the look
    panel.grid = element_blank(), # remove grid
    panel.background = element_blank(), # remove background
    axis.ticks.length.x = unit(1, "mm"), # set length of axis ticks on x axis
    axis.ticks.length.y = unit(1, "mm") # set length of axis ticks on y axis
  ) +

  labs( # label axes
    x = "Edge Distance (m)", # label x axis
    y = "Presence of Swift Parrots (yes/no)", # label y axis
    color = "Season", # color based on group, season
    fill = "Season" # fill color based on season
  )

mod_preds_plot # show the plot
```

## i. Write a caption for your figure

**Figure 1. Swift Parrots tend to occupy more nest boxes that are closer to the forests for seasons 2016 and 2019, and occupied more nest boxes in 2016.** Data from 'A case study of a critically endangered bird' (Stojanovic, Dejan et al. (2021). Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird [Dataset]. Dryad.). Points (blue) represent if Swift Parrots were present (yes = 1, no = 0) and how far the nest boxes were from the forest edge (m) (total n = 227). Lines with ribbons represent the model prediction of Swift Parrot nest box occupancy with a 95% confidence interval based on season (orange: 2016, purple: 2019).

## j.  Calculate model predictions

```{r}
mod_preds <- ggpredict(model2, # creating a new object for model predictions using model 2
                        terms = c("edge_distance [0:900] by = 1", "season")) |> # with the model, use predictors edge distance (0-900m) increasing by increments of 1 and season (2016 or 2019)
  rename(distance_to_edge = x, # rename x to distance_to_edge
         season = group) # rename group to season, makes understanding output easier

summary(mod_preds) # print summary of output

as.data.frame(mod_preds[mod_preds$distance_to_edge == 900, ]) # only show data for 900m
as.data.frame(mod_preds[mod_preds$distance_to_edge == 0, ]) # only show data for 0m

# makes it easier to see which data to focus on in interpretation
```

## k. Interpret your results

As the distance from the forest edge increases (m), the predicted probability of Swift Parrots occupying nest boxes decreased (Figure 1), showing that there is a negative relationship between probability of occupancy and forest edge distance. At the edge of the forest (0 m), the predicted probability of nest box occupancy was 0.48 (95% CI: [0.33, 0.64], $\alpha$ = 0.05) in 2016, and 0.30 (95% CI: [0.18, 0.44], $\alpha$ = 0.05) in 2019, demonstrating how Swift Parrots are more likely to occupy nest boxes near the forest edge, particularly in 2016. At 900 m away from the edge of the forest, the predicted probability of nest box occupancy went down to 0.12 (95% CI: [0.06, 0.24], $\alpha$ = 0.05) in 2016 and to 0.06 (95% CI: [0.03, 0.13], $\alpha$ = 0.05) in 2019, showing again a lower occupancy of nest boxes that are farther from the forest edge for both seasons. We conclude that Swift Parrots favor nest boxes that are closer to the edge of the forest because of the decreasing probability of occupancy as the distance increased from the forest edge, which could be because of resource availability and selective nesting preferences. The permanent boxes (those that remained in 2019) may have been taken over by competitor species, the Common Starling and the Tree Martin, because of the significantly lower Swift Parrot occupancy.

# Problem 4. Affective and exploratory visualizations

## a. Comparing visualizations

- My visualizations are different to each other because of their obvious visual contrast, where my affective visualization is creative and my exploratory visualization is very straight forward. My affective visualization involves more predictor variables (temperature, shoe type), instead just being on the phone or not.

- My visualizations are similar because they both show the median, first quantile, third quantile, and maximum and minimum.

- Because I had more data by the time I was working on homework 3, I have a lot more information in my affective visualization. I noticed between both that I tend to walk slower when I talk on the phone, and in my affective visualization, the type of shoe and temperature didn't have that much of an impact on the time I took. My median for the time I take to walk to class (when I am on the phone) dropped on my affective visualization compared to the exploratory, and there was a much bigger range from 18-23 instead of 20-21. My range also expanded for when I was not on the phone from 15-23 instead of 16-19. The exploratory visualization, even though there are limited data points, clearly shows that it took me longer to walk to class when I was on the phone, whereas the affective visualization shows a less obvious trend.

- I initially was only going to draw one shoe and add on both sets of data in different colors, but my peer reviewers (Ellie and Tanveer) suggested that I do two different shoes. They also said it looked a little busy with the pattern of the shoe so I decided to just have a solid background. They said it would be beneficial and more informative to add more variables, so I added an icon for each data point that represented my shoe type, and it's color represented the temperature that day. They agreed with what An suggested about incorporating the first and third quantiles in addition to the median, minimum, maximum. I did this by adding gaps in the shoe for each quantile (consistent with a clipart-esque design), and putting in dark lines for where the median, minimum, and maximum are.

## b. Sharing your affective visualization

I was present during week 10 of workshop.