---
title: "Final"
format: html
---

Link to GitHub repo: https://github.com/eliana427/ENVS-193DS_spring-2025_final.git
```{r}
library(tidyverse)
library(janitor)
sst <- read_csv("SST_update2023.csv")
```

Problem 1. Research writing

a. Transparent statistical methods

In part one, my co-worker likely did a Pearson correlation (if each variable is normally distributed), but if they're not normally distributed than the test was likely a Spearman rank correlation. In part 2, my co-worker likely did a one way ANOVA test (for normally distributed variable and sources have equal variances) or a Kurskal-Wallis test (data is not normally distributed).

b.  More information needed

The effect size (in this case, $\eta^2$) would be helpful because it gives information about how big the difference is between groups, specifically how much the type of source impacts the variation of average nitrogen load. A post hoc Tukey's HSD test would be helpful because it provides information about which groups in particular differ in nitrogen load by comparing the pairs of sources directly.

c. Suggestions for rewriting

We found a large difference ($\eta^2$ = effect size) in average load of nitrogen (kg/year) between tested sources (urban land, atmospheric deposition, fertilizer, waste water treatment, and grasslands) (one-way ANOVA, F(4, df within groups) = F value, p = 0.03, $\alpha$ = significance level)). A post hoc comparison using Tukey's HSD indicated that there were differences in the mean load of nitrogen between two of the of sources (amount kg/year less, x% CI: [lower bound, upper bound]) and another two of the of sources (amount less kg/year, x% CI: [lower bound, upper bound]).


Problem 2. Data visualization

a. Cleaning and summarizing
```{r}
library(lubridate)

sst_clean <- sst |> # store sst as new object called sst_clean
  clean_names() # make all names lowercase, and if there were spaces then change to underscores

colnames(sst_clean) # shows names of columns in data frame

sst_clean <- sst_clean |> # assigns sst_clean to itself to use in next functions
  mutate(
    date = as.Date(date), # make sure that R reads this column as a date that can be separated
    month = month(date, label = TRUE, abbr = FALSE), # separate out months from date column
    year_extracted = year(date) # extract the year from the data set
  ) |> 
  filter(year_extracted >= 2018) |> # filter the data set to only include dates 2018 and later
  mutate(year = factor(year_extracted), # read year as a factor
    date = NULL, # remove date column from displayed data in console
    site = NULL, # remove site column from displayed data in console
    latitude = NULL, # remove latitude column from displayed data in console
    longitude = NULL # remove longitude column from displayed data in console
  ) |> 
  group_by(year, month) |> # group by year and month
  summarize(mean_monthly_sst = mean(temp, na.rm = TRUE)) |> # calculate mean temperature per group
  ungroup() |>  # removes groupings back to original data frame

# rename all months in month column to abbreviations  
mutate(month = case_when(
  month == "January" ~ "Jan",
  month == "February" ~ "Feb",
  month == "March" ~ "Mar",
  month == "April" ~ "Apr",
  month == "May" ~ "May",
  month == "June" ~ "Jun",
  month == "July" ~ "Jul",
  month == "August" ~ "Aug",
  month == "September" ~ "Sep",
  month == "October" ~ "Oct",
  month == "November" ~ "Nov",
  month == "December" ~ "Dec"
)) |> 
  
mutate(month = factor(month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
ordered = TRUE)) # sort months in order when they appear on graph

slice_sample(sst_clean, n = 5) # display five random rows of the data set 

str(sst_clean) # show structure of the data set
```

b. Visualize the data

```{r}
sst_plot <- sst_clean # creating new object for plot

ggplot(data = sst_clean, # use data from sst_clean
       aes(x = month, # x axis is month
           y = mean_monthly_sst, # y axis is mean_monthly_sst
           color = year, # color will be based on year
           group = year)) + # data will be grouped by year+ 
geom_line() + # create a line plot
  scale_color_manual(values= c("#2E004F", "#4B0082", "#6A0DAD", "#8A2BE2", "#A24AC3", "#BF65D9")) + # colors will be a gradient with 2018 being the darkest and 2023 being the lightest
geom_point(size = 1.1) + # add points to the lines of size 1.1
  labs(x = "Month", # rename the x axis
       y = "Mean monthly sea surface temperature (°C)", # rename the y axis
       color = "Year") + # rename the legend which is based on color
  theme(panel.grid = element_blank(), # remove the grid lines
        panel.background = element_blank(), # make panel background blank
        panel.border = element_rect(color = "grey24", fill = NA, linewidth = 0.5), # add border to plot of width 0.5 but don't fill it in
        axis.ticks.length.x = unit (1, "mm"), # make axis ticks on x axis 1 mm
        axis.ticks.length.y = unit (1, "mm"), # make axis ticks on y axis 1 mm
        legend.position = "inside", # place legend inside of plot
        legend.position.inside = c(0.15, 0.7)) # legend will be positioned at these coordinates

```
